Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-07-06 write-ups/FSE/extended/paper.tex
Made paper cite extended version.

\bibliographystyle{abbrvnat}
\bibliography{tyler}

>>>>

\begin{thebibliography}{34}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Alglave et~al.(2015)Alglave, Batty, Donaldson, Gopalakrishnan, Ketema,
  Poetzl, Sorensen, and Wickerson]{ABDGKPSW-2015}
J.~Alglave, M.~Batty, A.~F. Donaldson, G.~Gopalakrishnan, J.~Ketema, D.~Poetzl,
  T.~Sorensen, and J.~Wickerson.
\newblock {GPU} concurrency: Weak behaviours and programming assumptions.
\newblock In \emph{ASPLOS}, pages 577--591. ACM, 2015.

\bibitem[Burtscher et~al.(2012)Burtscher, Nasre, and Pingali]{BNP12}
M.~Burtscher, R.~Nasre, and K.~Pingali.
\newblock A quantitative study of irregular programs on {GPU}s.
\newblock In \emph{IISWC}, pages 141--151. IEEE, 2012.

\bibitem[Cederman and Tsigas(2008)]{DBLP:conf/egh/CedermanT08}
D.~Cederman and P.~Tsigas.
\newblock On dynamic load balancing on graphics processors.
\newblock In \emph{EGGH}, pages 57--64, 2008.

\bibitem[Che et~al.(2013)Che, Beckmann, Reinhardt, and Skadron]{Pannotia}
S.~Che, B.~M. Beckmann, S.~K. Reinhardt, and K.~Skadron.
\newblock Pannotia: Understanding irregular {GPGPU} graph applications.
\newblock In \emph{IISWC}, pages 185--195, 2013.

\bibitem[Davidson et~al.(2014)Davidson, Baxter, Garland, and
  Owens]{DBLP:conf/ipps/DavidsonBGO14}
A.~A. Davidson, S.~Baxter, M.~Garland, and J.~D. Owens.
\newblock Work-efficient parallel {GPU} methods for single-source shortest
  paths.
\newblock In \emph{IPDPS}, pages 349--359, 2014.

\bibitem[{Google}()]{chrome-experiments}
{Google}.
\newblock {C}hrome {E}xperiments.
\newblock \url{https://www.chromeexperiments.com}.

\bibitem[Gupta et~al.(2012)Gupta, Stuart, and Owens]{owens-persistent}
K.~Gupta, J.~Stuart, and J.~D. Owens.
\newblock A study of persistent threads style {GPU} programming for {GPGPU}
  workloads.
\newblock In \emph{InPar}, pages 1--14, 2012.

\bibitem[Harish and Narayanan(2007)]{DBLP:conf/hipc/HarishN07}
P.~Harish and P.~J. Narayanan.
\newblock Accelerating large graph algorithms on the {GPU} using {CUDA}.
\newblock In \emph{HiPC}, pages 197--208, 2007.

\bibitem[Herlihy and Shavit(2008)]{HS08}
M.~Herlihy and N.~Shavit.
\newblock \emph{The Art of Multiprocessor Programming}.
\newblock Morgan Kaufmann Publishers Inc., 2008.

\bibitem[Howes()]{PersonalCommunicationHowes}
L.~W. Howes.
\newblock Personal communication.
\newblock Editor of the OpenCL 2.0 specification. 10 September 2016.

\bibitem[{HSA Foundation}(2016)]{HSAprogramming11}
{HSA Foundation}.
\newblock {HSAIL} virtual {ISA} and programming model, compiler writer, and
  object format {(BRIG)}, February 2016.
\newblock \url{http://www.hsafoundation.com/standards/}.

\bibitem[Kaleem et~al.(2016)Kaleem, Venkat, Pai, Hall, and
  Pingali]{DBLP:conf/ipps/KaleemVPHP16}
R.~Kaleem, A.~Venkat, S.~Pai, M.~W. Hall, and K.~Pingali.
\newblock Synchronization trade-offs in {GPU} implementations of graph
  algorithms.
\newblock In \emph{IPDPS}, pages 514--523, 2016.

\bibitem[{Khronos Group}(2015)]{opencl2Spec}
{Khronos Group}.
\newblock The {OpenCL} specification version: 2.0 (rev. 29), July 2015.
\newblock \url{https://www.khronos.org/registry/cl/specs/opencl-2.0.pdf}.

\bibitem[M{\'{e}}ndez{-}Lojo et~al.(2012)M{\'{e}}ndez{-}Lojo, Burtscher, and
  Pingali]{DBLP:conf/ppopp/Mendez-LojoBP12}
M.~M{\'{e}}ndez{-}Lojo, M.~Burtscher, and K.~Pingali.
\newblock A {GPU} implementation of inclusion-based points-to analysis.
\newblock In \emph{PPoPP}, pages 107--116, 2012.

\bibitem[Merrill et~al.(2015)Merrill, Garland, and
  Grimshaw]{DBLP:journals/topc/MerrillGG15}
D.~Merrill, M.~Garland, and A.~S. Grimshaw.
\newblock High-performance and scalable {GPU} graph traversal.
\newblock \emph{{TOPC}}, 1\penalty0 (2):\penalty0 14, 2015.

\bibitem[Muyan{-}{\"{O}}z{\c{c}}elik and
  Owens(2016)]{DBLP:conf/ppopp/Muyan-OzcelikO16}
P.~Muyan{-}{\"{O}}z{\c{c}}elik and J.~D. Owens.
\newblock Multitasking real-time embedded {GPU} computing tasks.
\newblock In \emph{PMAM}, pages 78--87, 2016.

\bibitem[Nobari et~al.(2012)Nobari, Cao, Karras, and
  Bressan]{DBLP:conf/ppopp/NobariCKB12}
S.~Nobari, T.~Cao, P.~Karras, and S.~Bressan.
\newblock Scalable parallel minimum spanning forest computation.
\newblock In \emph{PPoPPP}, pages 205--214, 2012.

\bibitem[{NVIDIA}(2016)]{PascalWhitepaper}
{NVIDIA}.
\newblock {NVIDIA Tesla P100}, 2016.
\newblock Whitepaper WP-08019-001\_v01.1.

\bibitem[\nvidia(2016)]{cuda-75}
\nvidia.
\newblock {CUDA} {C} programming guide, version 7.5, July 2016.

\bibitem[Pai and Pingali(2016)]{DBLP:conf/oopsla/PaiP16}
S.~Pai and K.~Pingali.
\newblock A compiler for throughput optimization of graph algorithms on {GPUs}.
\newblock In \emph{OOPSLA}, pages 1--19, 2016.

\bibitem[Park et~al.(2015)Park, Park, and Mahlke]{DBLP:conf/asplos/ParkPM15}
J.~J.~K. Park, Y.~Park, and S.~A. Mahlke.
\newblock Chimera: Collaborative preemption for multitasking on a shared {GPU}.
\newblock In \emph{ASPLOS}, pages 593--606, 2015.

\bibitem[Prabhu et~al.(2011)Prabhu, Ramalingam, Might, and
  Hall]{DBLP:conf/popl/PrabhuRMH11}
T.~Prabhu, S.~Ramalingam, M.~Might, and M.~W. Hall.
\newblock {EigenCFA}: accelerating flow analysis with {GPUs}.
\newblock In \emph{POPL}, pages 511--522, 2011.

\bibitem[Richards()]{PersonalCommunicationRichards}
A.~Richards.
\newblock Personal communication.
\newblock CEO of Codeplay Software Ltd. 2 September 2016.

\bibitem[{RISC OS}()]{risc-os-multitasking}
{RISC OS}.
\newblock Preemptive multitasking.
\newblock \url{http://www.riscos.info/index.php/Preemptive_multitasking}.

\bibitem[Smith and Anandtech(2016)]{anandtech}
R.~Smith and Anandtech.
\newblock Preemption improved: Fine-grained preemption for time-critical tasks,
  2016.
\newblock
  \url{http://www.anandtech.com/show/10325/the-nvidia-geforce-gtx-1080-and-1070-founders-edition-review/10}.

\bibitem[Solomon et~al.(2010)Solomon, Thulasiraman, and
  Thulasiram]{DBLP:conf/hpcc/SolomonTT10a}
S.~Solomon, P.~Thulasiraman, and R.~K. Thulasiram.
\newblock Exploiting parallelism in iterative irregular maxflow computations on
  {GPU} accelerators.
\newblock In \emph{HPCC}, pages 297--304, 2010.

\bibitem[Sorensen et~al.(2016)Sorensen, Donaldson, Batty, Gopalakrishnan, and
  Rakamaric]{DBLP:conf/oopsla/SorensenDBGR16}
T.~Sorensen, A.~F. Donaldson, M.~Batty, G.~Gopalakrishnan, and Z.~Rakamaric.
\newblock Portable inter-workgroup barrier synchronisation for {GPUs}.
\newblock In \emph{OOPSLA}, pages 39--58, 2016.

\bibitem[Tanasic et~al.(2014)Tanasic, Gelado, Cabezas, Ram{\'{\i}}rez, Navarro,
  and Valero]{DBLP:conf/isca/TanasicGCRNV14}
I.~Tanasic, I.~Gelado, J.~Cabezas, A.~Ram{\'{\i}}rez, N.~Navarro, and
  M.~Valero.
\newblock Enabling preemptive multiprogramming on {GPUs}.
\newblock In \emph{ISCA}, pages 193--204, 2014.

\bibitem[Tarpenning(1991)]{Tarpenning:1991:CMC:136810.136820}
M.~Tarpenning.
\newblock Cooperative multitasking in c++.
\newblock \emph{Dr. Dobb's J.}, 16\penalty0 (4), Apr. 1991.

\bibitem[Tzeng et~al.(2010)Tzeng, Patney, and Owens]{TPO10}
S.~Tzeng, A.~Patney, and J.~D. Owens.
\newblock Task management for irregular-parallel workloads on the {GPU}.
\newblock In \emph{HPG}, pages 29--37, 2010.

\bibitem[Vallina{-}Rodriguez and
  Crowcroft(2013)]{DBLP:journals/comsur/Vallina-RodriguezC13}
N.~Vallina{-}Rodriguez and J.~Crowcroft.
\newblock Energy management techniques in modern mobile handsets.
\newblock \emph{{IEEE} Communications Surveys and Tutorials}, 15\penalty0
  (1):\penalty0 179--198, 2013.

\bibitem[Vineet et~al.(2009)Vineet, Harish, Patidar, and
  Narayanan]{DBLP:conf/egh/VineetHPN09}
V.~Vineet, P.~Harish, S.~Patidar, and P.~J. Narayanan.
\newblock Fast minimum spanning tree for large graphs on the {GPU}.
\newblock In \emph{HPG}, pages 167--171, 2009.

\bibitem[Wang et~al.(2016)Wang, Davidson, Pan, Wu, Riffel, and
  Owens]{DBLP:conf/ppopp/WangDPWRO16}
Y.~Wang, A.~A. Davidson, Y.~Pan, Y.~Wu, A.~Riffel, and J.~D. Owens.
\newblock Gunrock: a high-performance graph processing library on the {GPU}.
\newblock In \emph{PPoPP}, pages 11:1--11:12, 2016.

\bibitem[Xiao and Feng(2010)]{XF10}
S.~Xiao and W.~Feng.
\newblock Inter-block {GPU} communication via fast barrier synchronization.
\newblock In \emph{IPDPS}, pages 1--12, 2010.

\end{thebibliography}

### 2017-07-06 write-ups/FSE/paper.tex
Made paper cite extended version.

We present our cooperative kernels programming model as an extension
to OpenCL.
%However, the cooperative kernels concept is more general,
%and could be applied to extend other GPU programming models, such as
%CUDA and HSA.
We describe the semantics of the model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}) and outline
important nonfunctional properties that the model requires to work
successfully (\mysec\ref{sec:nonfunctional}).
%and explain that the
%model is backwards-compatible
%(\mysec\ref{sec:backwardscompatibility}).
%

@@

%Among the GPUs available to us, four met our requirements: Intel HD
%520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
%drivers for OpenCL suffer from a known defect whereby long-running
%kernels lead to unkillable defunct
%processes~\cite{amd-defunct-process}.
%
%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.
%
\TSAdded{We thus ran our experiments on three Intel GPUs: HD 520, HD 5500
and Iris 6100. The results were similar across the GPUs, so for
conciseness, we report only on the Iris 6100 GPU (driver
20.19.15.4463) with a host CPU i3-5157U. The Iris has a reported 47
compute units. Results for the other Intel GPUs are available in
supplementary material, found on the author's website.}

@@

\myparagraph{Responsiveness}
%
\myfiglong\ref{fig:fine-grained-timing} reports, on three
configurations, the average gather and execution times for the
non-cooperative kernel with respect to the quantity of workgroups allocated to
it.  A logarithmic scale is used for time since gather times tend to
be much smaller than execution times. The horizontal grey lines
indicates the desired period for non-cooperative kernels.  These
graphs show a representative sample of our results; the full set of
graphs for all configurations is provided in supplementary material.

>>>>

We describe the semantics of our cooperative kernels programming model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}) and outline
nonfunctional properties required by the model (\mysec\ref{sec:nonfunctional}).
In the extended version of this paper we present a formal semantics, and discuss alternative semantic choices~\cite{ExtendedVersion}.
%and explain that the
%model is backwards-compatible
%(\mysec\ref{sec:backwardscompatibility}).
%

@@

%Among the GPUs available to us, four met our requirements: Intel HD
%520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
%drivers for OpenCL suffer from a known defect whereby long-running
%kernels lead to unkillable defunct
%processes~\cite{amd-defunct-process}.
%
%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.
%
\TSAdded{We thus ran our experiments on three Intel GPUs: HD 520, HD 5500
and Iris 6100. The results were similar across the GPUs, so for
conciseness, we report only on the Iris 6100 GPU (driver
20.19.15.4463) with a host CPU i3-5157U. The Iris has a reported 47
compute units. Results for the other Intel GPUs are available in
the extended version of this paper~\cite{ExtendedVersion}.}

@@

\myparagraph{Responsiveness}
%
\myfiglong\ref{fig:fine-grained-timing} reports, on three
configurations, the average gather and execution times for the
non-cooperative kernel with respect to the quantity of workgroups allocated to
it.  A logarithmic scale is used for time since gather times tend to
be much smaller than execution times. The horizontal grey lines
indicates the desired period for non-cooperative kernels.  These
graphs show a representative sample of our results; the full set of
graphs for all configurations is provided in the extended version of this paper~\cite{ExtendedVersion}.
